\section{CRDT Liftability} \label{sec:crdtliftability}

% \ap{A lot of complexity for this proof is hidden in the get function. Is there a way to characterise it in more detail?}

% \ps{Explain and illustrate the key definitions}


% \begin{definition}[Sequential Type]
% A sequential type $T = (S,s_0, \Delta)$  is a triple
%     where $S$ is the set of states of an object of type $T$, $s_0 \in S $ the initial state and $\Delta: S \times Op \to S$ is the transition function mapping a state and an operation to a successor state. 
    
%     We write $s \xrightarrow{op} s'$ when $\Delta(s, op) = s'$
% \end{definition}

% % \ps{I think you should start with the definition of liftability before characterizing it. This can be extracted from the proof below.}

% \begin{definition}[Induced binary operator]
%   Given a sequential type $T$, we define the binary operator $\oplus: S \times S \to S$ induced by the transition function $\Delta$ such that for any states $s , s', s'' \in S$, $s \oplus s' = s''$ by applying the union of the operation multi-sets that lead to states $s$ and $s'$ respectively. Formally, if $s_0 \xrightarrow{op_1 \dots  \op_m} s$ and $s_0 \xrightarrow{op_1' \dots  \op_n'} s'$. Then, $s \oplus s' = \Delta ^* (s_0, \{op_1 \dots op_m\} \uplus \{op_1'\dots op_n'\})$. $\Delta ^*$ is closure of operations, applied to the initial state $s_0$.


%   \ap{try to properly specify the confluence of the $\oplus$ operator as a property of the transition system. Relate this to intention preservation possibly. }

%   The binary operator $\oplus$ is well-defined only when the transition system is \emph{confluent}. 
%   A transition system is confluent if it satisfies the Church-Rosser theorem which states that the order in which operations are evaluated is irrelevant to the final value. 
%   This allows a transition system to reorder operations if the reordering does not cause the final value of a sequence of operations to change. \ps{define this term and discuss what does it implies.}

%   \ps{$\Delta^*$ is not defined.}
  
%   \ps{In this definition, operation $op_1'$ has to be applicable to the last state reached by $op_m$. This requires the transsition function to be total.}

%   \ps{What about when there are several sequences leading to the same state? Do we take a canonical one?}  

% \ap{The original definition of CRDT also defines a query operator in the tuple but never formally specifies its details. Do we include it here?}



% \end{definition}
 \begin{definition}[Liftablity]
A sequential type $T$ admits a $\R$-lifting if there exists a CRDT type $T\R = (\mathbb{S}, \bot,u, \sqcup)$ with an observation function $\get: \mathbb{S} \to S$ that materializes the value  of a CRDT and satisfies the following:

% and initial element $\bot$, and an observation function $\get : T\R \to T$, satisfying:

\ps{Recall that this is the definition of a state-based CRDT (i.e., a merge semi-lattice.)}

% \ap{need more explanation for L2. saying that merging the states does not loose any operations when using $\oplus$.}
\begin{itemize}
\item[(L1)] $\get(\bot) = s_0$
\item[(L2)] For any $s_1, s_2 \in \mathbb{S}, \get(s_1 \sqcup s_2) = \get(s_1) \oplus \get(s_2)$

% with $b$ fresh \ps{definition? One idea is that neither $a \leq b$, nor the other way around, but I see that you consider something different below.} with respect to $a$:


\item[(L3)] $\get$ is surjective: for every $s \in S$ there exists some $s' \in \mathbb{S}$ with $\get(s') = s$.
\item[(L4)] $\sqcup$ is commutative, associative, and idempotent.
\end{itemize}
 \end{definition}


%  \ap{Illustrate this definition with an example.}
% \ps{One thing also that matters is the assumption about $\oplus$. In general, a data type has a transition function. Maybe you can relate the two, e.g., by saying that $\oplus$ is induced by the transition function: for any two states $s$, $s'$ such that for some operation $op$ $op(s)=s'$ then $s \oplus s' = s'$.}

Consider the example of a CRDT presented in Example \ref{ex:counterCRDT}. 
A grow only counter is a lifting from a sequential counter without a decrement operation. 



\begin{theorem}[Liftability Characterization]
  \labtheo{lift}
  A sequential type $T$, with a set of states $S$ and its binary operator $\oplus$ is liftable via  $\R$ iff $(\oplus,S)$ is a commutative monoid and $\oplus$ is intention-preserving.
%   \ps{Well, this is not really $T$ but instead $(\oplus,S)$.}
\end{theorem}
This proof proceeds as follows: 
First, we prove the necessity condition i.e. Assuming that T\R ~exists, we show that T is a commutative monoid. 
Second, we prove sufficiency via a construction of a lifting for T\R.

\begin{proof}[Proof (Necessity): Liftability implies commutative monoid]$ $\newline

We verify the three axioms for $(S, \oplus)$ being a commutative monoid, namely associativity and commutativity of $\oplus$ and the existence of an identity element $s_0 \in S$. 
% \medskip
% \noindent\textbf{($\Rightarrow$) Necessity: Liftability implies commutative monoid}
% \medskip
% \noindent

% We verify the three axioms of a commutative monoid in turn.

\medskip
\noindent\emph{Associativity of $\oplus$.}
Let $s, t, u \in S$ be some arbitrary states. 
Since a valid lifting exists, by surjectivity (L3), choose states $a, b, c \in \mathbb{S}$ such that $\get(a) = s$, $\get(b) = t$, $\get(c) = u$.

% and where $a, b, c$ are mutually fresh with respect to one another (this is always achievable by considering distinct replica contributions).
% \ap{I need to characterize this freshness better. This reads very vague.}

By (L2) and associativity of $\sqcup$ (L4):
\begin{align}
\get(a \sqcup (b \sqcup c)) &= \get(a) \oplus \get(b \sqcup c) = s \oplus (t \oplus u) \label{eqn:assoc1}\\ 
\get((a \sqcup b) \sqcup c) &= \get(a \sqcup b) \oplus \get(c) = (s \oplus t) \oplus u \label{eqn:assoc2}
\end{align}

Since $\sqcup$ is associative, $a \sqcup (b \sqcup c) = (a \sqcup b) \sqcup c$. 
From equations \ref{eqn:assoc1} and \ref{eqn:assoc2}, we have: $s \oplus (t \oplus u) = (s \oplus t) \oplus u$. 
Since $s, t, u$ were arbitrary, $\oplus$ is associative.
% \ps{OK}


\medskip
\noindent\emph{Commutativity of $\oplus$.}
Let $s, t \in S$ be arbitrary states. By surjectivity (L3), choose $a, b \in \mathbb{S}$ with $\get(a) = s$ and $\get(b) = t$.

By (L2): $\get(a \sqcup b) = \get(a) \oplus \get(b) = s \oplus t$

% By commutativity of $\sqcup$ (L4): $a \sqcup b = b \sqcup a$

% Applying $\get$ and (L2) to the right hand side:
% \[
% \get(b \sqcup a) = \get(b) \oplus \get(a) = t \oplus s
% \]

Since $(a \sqcup b) = (b \sqcup a)$, we conclude:
\begin{align*}
 get(a \sqcup b) & = get(b \sqcup a) \\ 
 get(a) \oplus get(b) & = get(b) \oplus get(a) \\
 s \oplus t & = t \oplus s 
\end{align*}

Since $s, t$ were arbitrary, $\oplus$ is commutative. 


\medskip
\noindent\emph{Identity element.}
Let $t \in S$ be arbitrary and choose $a \in \mathbb{S}$ with $\get(a) = t$. 
We need a state representing $s_0$ to merge with. By (L1), $\get(\bot) = s_0$. 
$\bot$ is a state which carries no updates.
% We need $\bot$ to act as a fresh contribution with respect to $a$, which holds at initialization since $\bot$ carries no updates.

So, by (L2): $\get(a \sqcup \bot) = \get(a) \oplus \get(\bot) = t \oplus s_0$

Since $\bot$ is the identity of $\sqcup$ by the CRDT definition, so $a \sqcup \bot = a$. Thus, $\get(a \sqcup \bot) = \get(a) = t$. 
% Therefore $t \oplus s_0 = t$. 
Since $\oplus$ must be intention preserving, $s_0$ must not contain any updates so that $t \oplus s_0 = t$. 
Which makes $s_0 \in S$ the identity element for the commutative monoid $(\oplus, S)$. 

By commutativity of $\oplus$, $s_0$ is a two-sided identity for $\oplus$.
% \ps{OK}

This completes the proof of necessity.
\end{proof}

\begin{proof}[Proof (Sufficiency): Commutative monoid implies liftability]$ $\newline
  
 Assume that for a type $T$, its set of states $S$ and binary operator $\oplus$ form a commutative monoid. We construct a valid $T\R$-lifting explicitly.

\noindent\emph{Construction:} We define $T\R  = (\mathbb{S}, \bot, \Delta)$ as follows:
% \ps{I do not understand, $T\R$ should be a CRDT. So why using $\Delta'$ here?}

\begin{itemize}
\item \textbf{Initial element:} $\bot := s_0$, the constant function returning $s_0$.

\item \textbf{Merge operator:} $(f \sqcup g)(r) := f(r) \oplus g(r)$, applied pointwise.

\item \textbf{Observation function:} $\get(f) := \bigoplus_{r \in R} f(r)$, the fold of $\oplus$ over all replica values.
\end{itemize}

\ps{In what precedes, there is no definition of $\mathbb{S}$. From the definition of $\bot$ it should be a set of functions??}




\noindent\emph{Construction.} Let $R$ be a finite set of replica identifiers

\ps{One interesting (but not directly related) question is how necesary this assumption is. For instance, what happens is the systme is anonymous. This can be useful to bound metadata usage}. \ap{The replica identifiers here, simplify the get function w.r.t freshness constraints. I will try to find a way to eliminate this.}


Define:
\[
T\R := (R \to T)
\]
that is, the type of functions assigning a value in $T$ to each replica, which we think of as a state vector. Define:

\begin{itemize}
\item \textbf{Initial element:} $\bot := \lambda r.\, e$, the constant function returning $e$.

\item \textbf{Merge operator:} $(f \sqcup g)(r) := f(r) \oplus g(r)$, applied pointwise.

\item \textbf{Observation function:} $\get(f) := \bigoplus_{r \in R} f(r)$, the fold of $\oplus$ over all replica values.
\end{itemize}

Each replica $r$ maintains only its own coordinate, so replica $r$'s update is:
\[
\delta_r(f, t)(r') := \begin{cases} 
f(r') \oplus t & \text{if } r' = r \\
f(r') & \text{otherwise}
\end{cases}
\]

\ps{OK.}

\medskip
\noindent\emph{Verification of CRDT axioms.}

\noindent\textit{Commutativity of $\sqcup$:} For any $f, g$ and any $r$:
\[
(f \sqcup g)(r) = f(r) \oplus g(r) = g(r) \oplus f(r) = (g \sqcup f)(r)
\]
using commutativity of $\oplus$. Hence $f \sqcup g = g \sqcup f$.

\ps{We might need also that $e$ is neutral; no? This is easy: $r$, $(f \sqcup \bot)(r) = f(r) \oplus \bot(r) = f(r) \oplus e = f(r)$} \ap{The proof of e (now called $s_0$) being neutral comes from the other direction where we define the CRDT type and prove that it produces a commutative monoid.}

\noindent\textit{Associativity of $\sqcup$:} Follows immediately from associativity of $\oplus$ applied pointwise.

\noindent\textit{Idempotency of $\sqcup$:} This follows from the disjoint union in the definition of $\oplus$. Since merge is performed pointwise, every operation is only counted once in the binary operator. Thus, remaining idempotent.  


% This requires an additional condition. Since $\oplus$ need not be idempotent in general, we enforce idempotency structurally by the freshness discipline. Each replica update $\text{inc}_r$ is applied at most once per value, so the case $f \sqcup f$ never arises in practice. Alternatively, one can replace $T$ with the lifted multiset construction $T\R := (R \to \mathbb{N}) \otimes T$ and recover idempotency formally, at the cost of a more complex $\get$. \ps{Not sure to follow this last part.}

\medskip
\noindent\emph{Verification of the homomorphism condition (L2).}
Let $f, g \in T\R$ be fresh with respect to one another, meaning their nonzero coordinates are disjoint: for each $r$, at most one of $f(r)$, $g(r)$ is non-initial \ps{I see.}. Then:
\[
\get(f \sqcup g) = \bigoplus_{r \in R} (f \sqcup g)(r) = \bigoplus_{r \in R} (f(r) \oplus g(r))
\]

Since $\oplus$ is commutative and associative, and the supports of $f$ and $g$ are disjoint:
\[
= \left(\bigoplus_{r \in R} f(r)\right) \oplus \left(\bigoplus_{r \in R} g(r)\right) = \get(f) \oplus \get(g)
\]

\medskip
\noindent\emph{Verification of (L1).}
\[
\get(\bot) = \bigoplus_{r \in R} e = e
\]
using the identity law of $\oplus$ iterated over $R$.

\medskip
\noindent\emph{Verification of (L3).}
For any $t \in T$, fix any replica $r_0 \in R$ and define $f := \lambda r.\, \text{if } r = r_0 \text{ then } t \text{ else } e$. Then $\get(f) = t \oplus e \oplus \cdots \oplus e = t$. So $\get$ is surjective. \ps{OK}

\medskip
This completes the construction and verification. Hence $(T, \oplus, e)$ admits a $\R$-lifting.
\end{proof}

\medskip
\noindent\textbf{Conclusion}

$(T, \oplus, e)$ admits a $\R$-lifting if and only if $(T, \oplus, e)$ is a commutative monoid.
\hfill$\blacksquare$

%% \ap{present some existing CRDTs and show how this theorem applies to sequential types to get a CRDT.}
%% \ap{Further extends to the fact that the product of two comm monoids is a monoid. }

% \begin{corollary}[Non-liftability of string concatenation]
% The type $(\Sigma^*, \cdot, \varepsilon)$ admits no $\R$-lifting, since string concatenation is not commutative: $\text{``AB''} \cdot \text{``CD''} = \text{``ABCD''} \neq \text{``CDAB''} = \text{``CD''} \cdot \text{``AB''}$. By the necessity direction of the theorem, no valid $\get$ can exist. \qed
% \end{corollary}

\medskip
The result above provides some interesting observations about CRDTs.

\paragraph{Observations}
First, the fact that $(\mathbb{N},\mathit{max})$ is a commutative monoid implies the existence of a grow-only CRDT counter (see~\refsec{background}).
%
For a CRDT sequence, its states are all the partial orders $\mathcal{P}$ such that for any two partial orders $P_1=(V_0, \le_0)$ and $P_2=(V_1, \le_1)$, if (i) if $y \le_0 x$ and $x \in V_1$, then $y \le_1 x$, and (ii) the converse holds for $P_2$ wrt. $P_1$.
In that case, $(\mathcal{P}, \cup)$ where $\cup$ is the standard union operator forms a commutative monoid.

A second observation is that the Cartesian product of two commutative monoids is also a commutative monoid.
Thus, for any object $T$ with a set of states $S$ and mutators $\Delta_m$ if $(S,\Delta_m)$ can be split into a set $(S,\Delta_1), \ldots, (S,\Delta_m)$ such that each $(S,\Delta_i)$ is liftable in the sense of \reftheo{lift}, then $T$ admits a CRDT implemenation.
%
This happens with PN-counter (\refsec{background});
in this case $m=2$, $\Delta_1$ encodes increments, and $\Delta_2$ encodes decrements.

Going back to the CRDT sequence, removal is implemented by taking the Cartesian product of $(\mathcal{P}, \cup)$ together with a commutative monoid $(\mathcal{M}, \land)$ where every $m \in \mathcal{M}$ is a map from the sequence domain to $\{0,1\}$ such that $(m \land m')[x]$ equals $0$ if either $m[x]=0$ or $m'[x]=0$, and $1$ otherwise.
This decomposition correspond to the notion of tombstone, as found e.g., in WOOT~\cite{woot}.

A third observation is that there exists a universal approach to lift sequential objects:
Let $M \subseteq O$ be the mutators defined by the data type and $<$ be the common partial order over sequences in the free monoid $M^*$.
Suppose some linearization $\ll$ of $<$.
%
Then, denoting $\sqcup_{\ll}$ the lower upper bound operator induced by $\ll$, $(M^*,\sqcup_{\ll})$ is a commutative monoid.

Interestingly, this constructs differs from the standard last-write-wins (LWW) approach.
The next section proves that in fact LWW and other timestamps based approach to arbitration are not lifting a sequential data type.
